---
title: "Local Score Analysis for GWAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Local Score Analysis for GWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 7,
    fig.height = 4
)
```

## Why local scores?

Standard GWAS tests each marker independently and applies a FDR correction on
resulting p-values, e.g. by Benjamini-Hochberg or Bonferroni methods.  While
this controls for false positive discoveries, nearby markers are treated as
independent. This is rarely the case, as markers in linkage disequilibrium (LD)
carry correlated signal. This over-correction can lead to
moderately-significant associations being lost. 

The "local score" method (Fariello *et al.* 2017; Bonhomme et al 2019)
addresses this by aggregating accumulated signal over markers within a local
genome window. In simple terms, it compares the integrated signal over a run of
at least weakly-associated markers to a distribution of such windows under the
null (using Gumbel extreme-value theory), thereby distinguishing assocations as
longer than expected windows of higher than expected association. In doing so,
we detect associations even if no individual marker passes the global FDR
threshold.


## Basic usage

```{r load}
library(gwashelpers)
```

We simulate a chromosome with 500 markers.  Most markers are non-associated
(p-values drawn from a roughly uniform null), but positions 240–260 harbour a
genuine signal with very small p-values.

```{r simulate}
set.seed(42)
n = 500

# Null markers: uniform p-values, mildly autocorrelated to mimic LD
p = numeric(n)
p[1] = runif(1)
for (i in 2:n) p[i] = 0.2 * p[i - 1] + 0.8 * runif(1)

# Inject a signal cluster at positions 240-260
p[240:260] = runif(21, min = 1e-6, max = 5e-4)
```

Run `local_score()` to obtain per-marker scores, Lindley process values,
the genome-wide threshold, and window labels in a single data frame:

```{r run}
result = local_score(p, xi = 2)
head(result)
```

The four columns are:

| Column | Meaning |
|---|---|
| `score` | $X_m = -\log_{10}(p_m) - \xi$ |
| `lindley_score` | Lindley process value $h_m \ge 0$ |
| `lindley_thresh` | Genome-wide threshold $t$ (constant across rows) |
| `lindley_window` | Integer label for each excursion above $t$; `NA` below |

The threshold is shared by all markers because it depends only on the number
of markers and their average autocorrelation, not on position:

```{r thresh}
unique(result$lindley_thresh)
```


## Visualising the result

```{r plot, fig.cap = "Lindley process along a simulated chromosome. The dashed line is the genome-wide threshold; shaded regions are significant windows."}
thresh = unique(result$lindley_thresh)
pos    = seq_along(result$lindley_score)

# Base plot
plot(pos, result$lindley_score, type = "l", col = "steelblue",
     xlab = "Marker position", ylab = "Lindley score",
     main = "Local score across simulated chromosome")
abline(h = thresh, lty = 2, col = "firebrick", lwd = 1.5)

# Shade significant windows
in_window = !is.na(result$lindley_window)
if (any(in_window)) {
    runs = rle(in_window)
    ends   = cumsum(runs$lengths)
    starts = ends - runs$lengths + 1
    for (k in which(runs$values)) {
        rect(starts[k] - 0.5, 0, ends[k] + 0.5, max(result$lindley_score),
             col = adjustcolor("gold", alpha.f = 0.3), border = NA)
    }
    # Re-draw line on top of shading
    lines(pos, result$lindley_score, col = "steelblue")
    abline(h = thresh, lty = 2, col = "firebrick", lwd = 1.5)
}

legend("topleft", legend = c("Lindley score", "Threshold", "Significant window"),
       col = c("steelblue", "firebrick", adjustcolor("gold", alpha.f = 0.6)),
       lty = c(1, 2, NA), pch = c(NA, NA, 15), bty = "n")
```


## Extracting significant peaks

The `lindley_window` column labels each contiguous run of markers above the
threshold with a distinct integer.  Split the result to summarise each peak:

```{r peaks}
sig = result[!is.na(result$lindley_window), ]

if (nrow(sig) > 0) {
    peaks = do.call(rbind, lapply(split(sig, sig$lindley_window), function(w) {
        data.frame(
            window     = w$lindley_window[1],
            start      = which(result$lindley_window == w$lindley_window[1])[1],
            end        = tail(which(result$lindley_window == w$lindley_window[1]), 1),
            peak_score = max(w$lindley_score),
            n_markers  = nrow(w)
        )
    }))
    print(peaks, row.names = FALSE)
} else {
    message("No markers exceeded the threshold.")
}
```

The peak score is the maximum $h_m$ within the window.  All markers in the
window share the same integer label, so downstream code can group them by
`lindley_window` to retrieve the best-candidate SNP or to annotate the
associated genomic region.


## Choosing xi

The parameter `xi` controls which markers contribute positively: only signals
with p values below $10^{-{\xi}}$ contribute to the signal. A larger `xi`
ignores moderately-significant markers and demands stronger individual signal
before a region counts as associated.  The threshold is recomputed for each
`xi` value, so both the score and the significance criterion change together.

```{r xi_comparison}
r1 = local_score(p, xi = 1)
r2 = local_score(p, xi = 2)

cat(sprintf("xi=1: threshold = %.2f, peak Lindley score = %.2f\n",
            unique(r1$lindley_thresh), max(r1$lindley_score)))
cat(sprintf("xi=2: threshold = %.2f, peak Lindley score = %.2f\n",
            unique(r2$lindley_thresh), max(r2$lindley_score)))
```


## Reference

Fariello, M.-I. *et al.* (2017). Accounting for linkage disequilibrium in
genome scans for selection without individual genotypes: The local score
approach. *Molecular Ecology*, <https://doi.org/10.1111/mec.14141>

Bonhomme, M. *et al.* (2019) A local score approach improves
GWAS resolution and detects minor QTL: application to Medicago truncatula
quantitative disease resistance to multiple Aphanomyces euteiches isolates.
*Heredity* <https://doi.org/10.1038/s41437-019-0235-x>
